Step by Step: Using Step Selection Analysis to Simulate Dispersal and Assess Landscape Connectivity

### 2021-04-30 Chapter_1/04_Manuscript/Manuscript.tex
Worked on manuscript

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
For many species, a sufficient degree of landscape connectivity is a crucial
prerequisite for long-term species persistence. This is especially true for
animals that exhibit long-distance dispersal, as dispersers typically cross a
variety of habitats and potentially move into unfamiliar territory. Estimating
habitat permeability and quantifying landscape connectivity have therefore
become two major tasks in conservation ecology. For this, least-cost analysis
and circuit theory have been the two workhorses during the past two decades.
This is largely owed to their ease of use and intuitive nature which has
facilitated the application of the methods across a broad range of the animal
kingdom. However, both methods make several restricting assumptions that limit
their usefulness in reality.

Recent innovations in movement ecology, particularly on step selection analysis,
have brought forward novel ways to look at connectivity. For instance,
integrated step selection functions provide a means to parametrize a fully
mechanistic movement model based on which virtual dispersers could be simulated.
While this approach has been used to infer habitat utilization, it may also
serve to highlight landscape connectivity and pinpoint dispersal corridors.

Here, we propose the simulation of dispersal trajectories as a much more generic
way to identify dispersal barriers and potential movement corridors. To achieve
this, we applied integrated step selection analysis and uses data from 16
dispersing wild dogs to parametrize a fully mechanistic movement model for
dispersing African wild dogs. The model rendered dispersers' habitat and
movement preferences, as well as potential interactions between them. Based on
the model, we simulated myriads of dispersal trajectories and inferred dispersal
corridors across the earth's largest transboundary conservation area, the
Kavango-Zambezi Transfrontier Conservation Area. We also exemplify how such
simulations could be analysed using network theory. Finally, we discuss the
benefits and pitfalls of dispersal simualtions and highlight potential
improvements to be made in the future.
\end{abstract}

@@

% Change page numbering
\pagenumbering{arabic}

@@

% Temporal scale
Neither least-cost analyes or circuit theory are capable of rendering the
temporal dimension of movement. Such influences could, however, be studied using
individual based movement models. Still, only few have explored these
possibilities \citep{Kanangaraj.2013, Hauenstein.2019, Zeller.2020}.

@@

% Integrated step selection analysis
An animals movement trajectory can be seen as the result of an interplay between
habitat and movement preferences. Traditional step selection functions cannot
account for a correlation between turning angles and step lengths, unless the
two are sampled jointly from a copula distribution \citep{Hodel.2021}. Recently,
however, SSFs have been generalized to \textit{integrated} SSFs (iSSFs), which
tear apart movement and habitat preferences of the studied animals
\citep{Avgar.2016}. The method thus allows to render an animals preferences with
respect to prevailing habitat conditions, as well as with respect to movement
characteristics. This reduces potential biases in estimated preferences and
enables parametrization of a fully mechanistic movement model from which
movement and space use can be simulated \citep{Avgar.2016, Signer.2017}.
\cite{Signer.2017} used integrated step selection analysis to simulate steady
state utilization distributions of resident animals. However, the degree to
which such simulations could be used to simulate dispersers and infer movement
corridors is unknown.

@@

\section{Methods}
\subsection{Study Area}
The study area was centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) and was represented by a rectangular bounding box that
stretched over 1.3 Mio. km\textsuperscript{2}. Its extent encompassed the entire
KAZA-TFCA (\Cref{StudyArea}b), which is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, covering a total of 520'000 km\textsuperscript{2}. Its landscape
varies regionally and ranges from savanna, to grassland, and from dry to moist
woodland habitats. A dominant hydrogeographical feature in its center is the
flood-pulsing Okavango Delta, the earth's largest inland delta. The delta and
its surroundings are considered a stronghold for African wild dogs and may act
as a source for the recolonization of surrounding habitats \citep{Cozzi.2013}.
The wet season within the study area lasts from November to March and is out of
phase with the main flooding of the Okavango Delta which peaks between July and
August \citep{McNutt.1996, Wolski.2017}. While large portions of land within the
KAZA-TFCA are designated protected areas or national parks that remain largely
free from humans, substantial human influence originates from roads,
agricultural sites and several small to large settlements.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{(a) The study area of our research was confined by a bounding box
    encompassing the entire KAZA-TFCA, which comprises parts of Angola, Namibia,
    Botswana, Zimbabwe, and Zambia. (b) The KAZA-TFCA represents the world's
    largest terrestrial conservation area and covers a total of 520'000
    km\textsuperscript{2}. It will connect or reconnect multiple already
    existing national parks and protected areas (green polygons) that are
    distributed across its extent. The African wild dog dispersal data
    considered in this study was collected on free-ranging individuals departing
    from the Moremi National Park in northern Botswana (since game reserves in
    Botswana serve the same purpose as national parks, we refer to them as
    national parks for simplicity).}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2019, we collected GPS relocation data from dispersing wild
dogs in a free-ranging wild dog population inhabiting the Moremi National Park
in northern Botswana. We identified potential dispersers based on age, number of
same‐sex siblings, pack size, and presence of unrelated opposite-sex individuals
in the pack \citep{McNutt.1996, Behr.2020}. We immobilized individuals using a
cocktail of Ketamine/Xylazine/Atropine \citep{Osofsky.1996, Cozzi.2020},
injected with a dart, fired from a CO\textsubscript{2}-pressurized gun
(\textit{DAN-Inject, Denmark}). Anesthesia protocols were approved by the
Ministry of Environment, Natural Resources Conservation and Tourism of Botswana
(permit EWT 8/36/4 XXXVI). After immobilization, individuals were fitted with
GPS/Satellite radio collars (\textit{Vertex Lite; Vectronic Aerospace GmbH,
Berlin}) that included an automated drop-off mechanism. Handling and collaring
of all individuals was carried out and supervised by a Botswana-registered
wildlife veterinarian. All of the immobilized individuals usually rejoined their
pack within one hour after the procedure. Out of all collared individuals, 16
individuals eventually dispersed in separate same-sex coalitions and their
trajectories were successfully recorded (7 female and 9 male coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Collected relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. To distinguish between periods of
residency and dispersal, we applied the net-squared displacement metric to the
observed movement data. This metric measures the squared Euclidean distance of a
collared individual to a reference point \citep{Borger.2012}. As a reference
point, we used the center of each disperser's natal home range, such that
dispersal was deemed to have started when an individual left its natal home
range and ended once individuals became sedentary again. For the purpose of this
study, we discarded all data that was collected during residency and only
retained movement data during dispersal. Previous research suggests that females
and males behave similarly during dispersal \citep{Woodroffe.2019, Cozzi.2020},
so we did not distinguish between sexes in our analyses. Ultimately, we
converted the collected GPS coordinates to steps, where each step represented
the straight-line distance traveled by and individual between two consecutive
GPS relocations \citep{Turchin.1998}.

\subsection{Covariates}
We represented the physical landscape in the study area using a set of habitat
covariates rendering water-cover, distance to water, tree-cover, and
shrub/grassland-cover. Because water cover greatly changes within and between
years around the Okavango Delta, we applied a remote sensing technique that
allowed us to generate frequently updated water cover layers and corresponding
distance to water layers (see \citealp{Wolski.2017} and \citealp{Hofmann.2021}).
Hence, the obtained layers temporally aligned with our dispersal events.
Furthermore, we computed a proxy for human influence, depicting anthropogenic
pressures stemming from human-density, agricultural sites, and roads. We
coarsened or interpolated all layers to a target resolution of 250 m by 250 m
and cropped them to the extent of the KAZA-TFCA. Further details on the sources
and preparation of each habitat covariate are given in \cite{Hofmann.2020}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}). Because wild dogs follow a diurnal activity pattern, we also
coded a binary variable (\textsf{LowActivity}) indicating whether a step was
realized during low activity (17:00 to 07:00 local time) or high activity (09:00
to 17:00 local time). We conducted all cleaning and analysis, as well as all
simulations using the statistical software {\tt R}, version 3.6.6
\citep{R.2019}. The corresponding {\tt R}-scripts are provided on Github.

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In contrast to regular step selection
functions, \textit{integrated} step selection functions allow simultaneous
inference on movement and habitat preferences of the studied animal, and on
potential interactions between habitat and movement preferences. In result, the
method produces less biased selection estimates and allows to treat the
resulting model as a fully mechanistic movement model based on which movement
can be simulated. To conduct iSSF analysis, we paired each observed step with 24
random steps, alltogether forming a stratum that received a unique identifier.
Each of the 24 random steps served as pseudeo-absence and resembled a potential
alternative that the animal could have realized but decided not to. We generated
random steps by sampling random turning angles from a uniform distribution
(\(-\pi, +\pi\)) and step lengths from a gamma distirbution that was fitted to
observed steps (scale = 6'308, shape = 0.37). While the number of random steps
is inversely proportional to the sampling error \citep{Avgar.2016}, a
preliminary analysis revealed only minor differences in model performance when
sampling additional random steps, hence we deemed 24 random steps to be
sufficient. Along each step, we extracted and averaged spatial covariates using
the {\tt velox} package and we calculated the movement metrics \textsf{sl},
\textsf{log(sl)}, and \textsf{cos(ta)}. To facilitate model convergence, we
standardized all continuous covariates to a mean of zero and a standard
deviation of one. Since correlation among covariates was low (\(|r| > 0.6\);
\citealp{Latham.2011}), we retained all of them for modeling.

To compare realized steps (scored 1) to random steps (scored 2), we assumed that
animals assigned a selection score \(w(x)\) to each step. The score depended on
the step's covariates \(X\), as well as the animals preferences \(\beta\)
towards these covariates. The probability of a step being realized was then
contingent on the step's selection score, as well as on the selection scores of
all other step in the same stratum. To estimate the preferences of interest, we
fitted conditional logistic regression models in the r-package {\tt glmmTMB}
following the method proposed by \citep{Muff.2020}. This approach enables to fit
random slopes in addition to random intercepts. Here, we used animal IDs as
random effect and modelled random slopes for each covariate. To apply the
``poisson trick'', we fixed the random intercept variance at an arbitrary high
value of 10\textsuperscript{6} \citep{Muff.2020}.

The structure of our movement model was based on a habitat selection model for
dispersing wild dogs presented in \cite{Hofmann.2021}. In the original model, no
interactions between the habitat and movement covariates were considered, so we
slightly expanded this base model by proposing interactions between movement and
habitat covariates. More specifically, we started with the base model and
incrementally increased model complexity by adding all possible two-way
interactions between habitat covariates and movement covariates. For instance,
for the covariate \textsf{water}, we proposed the interactions
\textsf{Water:log(sl)}, \textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}.
Besides those combinations, we also proposed the interactions
\textsf{sl:cos(ta)} and \textsf{log(sl):cos(ta)} to account for a correlation
between turning angles and step lengths, as well as the interactions
\textsf{sl:LowActivity} and \textsf{log(sl):LowActivity} to account for the fact
that step lengths may differ due to wild dogs' diurnal activity pattern. To
compare competing models and assess the most parsimonious movement model, we ran
stepwise forward model selection based on Akaike's Information Criterion (AIC,
\citealp{Burnham.2002}) and computed model weights.

We validated the predictive power of the selected movement model using a k-fold
cross-validation procedure for case-control studies presented in
\cite{Fortin.2009}. We randomly split the input data into training and testing
data with an 80:20 ratio. Using the training data we parametrized a movement
model based on which we predicted selection scores \(w(x)\) for all steps in the
test data. Within each stratum we then assigned ranks 1-25 to each step based on
predicted selection scores such that rank 1 was given to the step with the
highest score \(w(x)\). Across all strata we determined the realized step's rank
and calculated rank frequencies for realized steps. Finally, we computed
Spearman's rank correlation between ranks and associated frequencies \(r_{s,
realized}\). We replicated the entire procedure 100 times and computed the mean
correlation coefficient (\(\bar{r}_{s, realized}\)), as well as its 95\%
confidence interval across all replicates. For comparison, we repeated the same
procedure 100 times assuming random preferences. Random preferences were
implemented by discarding the realized step from all strata and identifying the
rank of a random step in each stratum. Again, we calculated Spearman's rank
correlation coefficient (\(r_{s, random}\)), its mean across repetitions
(\(\bar{r}_{s, random}\)), and its 95\% confidence interval. This validation
proofs a significant prediction in case the confidence intervals of
\(\bar{r}_{s, realized}\) and \(\bar{r}_{s, random}\) do not overlap.

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate a total of 80'000
dispersing wild dogs moving across the KAZA-TFCA. The simulation resembled an
inverted iSSF and was set up as follows. (1) We sampled a source point at which
a disperser was initiated. We assumed a random initial orientation of the animal
in order to permit calculation of relative turning angles. (2) We generated a
set of 25 random steps by sampling turning angles from a uniform distribution
(\(-\pi, +\pi\)) and step lengths from the fitted gamma distribution. In line
with our training data, each step resembled the straight line movement realized
within 4 hours. To prevent the sampling of unreasonably large steps, we capped
sampled step lengths to a maximum of 35 km, which corresponds to the farthest
distance ever traveled within 4 hours by our dispersers. (3) Along each random
step, we extracted habitat covariates and calculated movement covariates. (4) We
applied the parameterized movement model to predict the selection score \(w(x)\)
for each of the proposed steps and translated selection scores into
probabilities by applying \Cref{EQ2}. (5) We sampled one of the random steps
based on predicted probabilities and used it to determine the animal's new
position. We then repeated steps (2) to (5) until 2'000 steps were realized.

Because some simulated dispersers would inevitably approach a map boundary, some
of the proposed random steps may leave the study area, such that covariates
could be extracted and no selection score could be calculated. To mitigate the
influence of such edge effects, we followed \citep{Koen.2010} and artificially
expanded all covariate layers by adding a 100 km buffer zone within which we
randomized covariate values by resampling values from the observed covariate
layers. This allowed simulated dispersers to leave and reenter the study area
through this buffer zone. Nevertheless, if a disperser approached this outer
boundary and one of the proposed random steps left the buffer, we resampled
transgressing steps until they fully lied within the buffer zone, thereby
forcing dispersers to bounce off the outer boundary. The simulation was mainly
coded in the programming language {\tt R}, yet a few helper functions were
written in {\tt C++} and imported into {\tt R} using the {\tt Rcpp} package
\citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Source Points}
To initiate virtual dispersers at biologically meaningful source points, we
randomly placed source points within contiguous protected areas larger \(>\) 700
km\textsuperscript{2} (\Cref{SourcePoints} (a)). This conforms to the average
home range requirement of resident wild dogs \citep{Pomilia.2015} and allowed us
to remove areas too small to host viable wild dog populations. By distributing
source points randomly, the number of source points per km\textsuperscript{2}
within protected areas was approximately equal across our study area. Across all
protected areas, we distributed 50'000 source points, each representing the
starting point of a single dispersal trajectory. To render potential immigrants
into the study system, we placed additional 30'000 source points inside the
buffer zone around the main study area (\Cref{SourcePoints} (b)).

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourceAreas.png}
    \caption{(a) Illustration of an artificially expanded covariate layer (water
    cover). We expanded the layer by adding a 100 km buffer (light blue) which
    we filled with values sampled from the original layer. (b) Different source
    areas from which we released virtual dispersers. We initiated 50'000
    dispersers within the main study area (green) and another 30'000 dispersers
    within a virtual buffer (blue).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Heatmaps}
To identify dispersal hotspots across our study area, we rasterized all
simulated trajectories and created a heatmap indicating the absolute frequency
at which each raster-cell in the study area was visited by our virtual
dispersers. If the same trajectory crossed a raster-cell twice, it was only
counted once. That is, we did not consider revisits, thereby mitigating biases
arising from individuals that were trapped in a certain locations and moved
around in circles. We achieved high performance rasterization of all simulated
trajectories using the R-package {\tt terra} \citep{Hijmans.2020}. While one
could easily subset the data to visualize heatmaps assuming different dispersal
durations, we retained all 2'000 simulated steps to create a single heatmap.

% To examine if and how ``heat'' changes in response to changes in the location of
% source points and the number of simulated steps, we followed a 2 x 6 design and
% created heatmaps for both point sampling regines, as well as for 68, 125, 250,
% 500, 1000, and 2000 dispersal steps. We quantified the similarity of the
% resulting 12 heatmaps to the permeability and least-cost corridor maps presented
% in (Hofmann ...) we used Bhattacharyya's affinity. Bhattacharyya's affinity
% ranges from zero (complete separation) to one (perfect match) and has earlier
% been proposed to compare the overlap of utilisation distributions (Fieberg).

>>>>

%------------------------------------------------------------------------------
%	Abstract
%------------------------------------------------------------------------------
\newpage
\begin{abstract}
For many species, dispersal represents a crucial process that allows to avoid
inbreeding, to colonize new habitats and to reinforce small and unviable
subpopulations. Successful dispersal is therefore a crucial pre-requisite for
long-term species persistence in wild animal populations. However, the ability
to disperse is contingent a sufficient degree of landscape connectivity, which
is why the estimation of connectivity and identification of dispersal corridors
has become a task of extraordinary importance in ecological studies and for the
planning of conservation initiatives.

Over the past two decades, scientists have mainly relied on analytical tools
such as least-cost methods and circuit theory to investigate landscape
connectivity. This is largely owed to their ease of use and intuitive nature,
which have facilitated their application in the past. Despite their usefulness
for various ecological applications, both methods make several restricting
assumptions that can limit their applicability in reality. Least-cost analysis,
for instance, suffers from the fact that animals are assumed to have perfect
knowledge about the entire landscape and choose cost-minimizing routes
accordingly. On the contrary, circuit theory assumes that animals walk according
to a complete random walk, which is equally unlikely to hold in reality.

Fortunately, recent innovations in movement ecology have brought forward novel
opportunities to study dispersal and landscape connectivity. In particular, the
rich suite of resource selection functions has undergone substantial
improvements in recent years. Most notably, step selection functions have been
generalized to \textit{integrated} step selection functions, which essentially
represent fully mechanistic movement models based on which individual animal
movement could be predicted and simulated. While such models have been applied
to simualte the steady state utilization distribution, similar approaches could
be useful for investigating transient movement behavior and study landscape
connectivity in a much more natural way.

Here, we showcase the use of integrated step selection analysis to simulate
dispersal of a top-predator across a large study area. For this, we collected
data on 16 dispersing wild dogs and analysed it in an integrated step selection
analysis to produce a fully mechanistic movement model rendering wild dog
dispersal. Based on this model, we simulated 80'000 dispersers moving across the
Kavango-Zambezi Transfrontier Conservation Area, the world's largest
transboundary conservation area. We summarize the simulation output in a heatmap
indicating the intensity of use across the KAZA-TFCA, and we apply network
theory to produce a set of complementary maps that further reveal major
dispersal hotspots and crucial bottlenecks. Finally, we discuss the benefits and
pitfalls of dispersal simualtions and we highlight improvements to be made in
the future.
\end{abstract}

@@

\onehalfspacing
\tableofcontents
\doublespacing

% Change page numbering
\newpage
\pagenumbering{arabic}

@@

% Temporal scale
Neither least-cost analyes or circuit theory are capable of rendering the
temporal dimension of movement. Such influences could, however, be studied using
individual based movement models. Still, only few have explored these
possibilities \citep{Kanagaraj.2013, Hauenstein.2019, Zeller.2020}.

@@

% Integrated step selection analysis
An animals movement trajectory can be viewed as the result of an interplay
between habitat and movement preferences. Traditional step selection functions
cannot account for a correlation between turning angles and step lengths, unless
the two are sampled jointly from a copula distribution \citep{Hodel.2021}.
Recently, however, SSFs have been generalized to \textit{integrated} SSFs
(iSSFs), which tear apart movement and habitat preferences of the studied
animals \citep{Avgar.2016}. The method thus allows to render an animals
preferences with respect to prevailing habitat conditions, as well as with
respect to movement characteristics. This reduces potential biases in estimated
preferences and enables parametrization of a fully mechanistic movement model
from which movement and space use can be simulated \citep{Avgar.2016,
Signer.2017}. \cite{Signer.2017} used integrated step selection analysis to
simulate steady state utilization distributions of resident animals. However,
the degree to which such simulations could be used to simulate dispersers and
infer movement corridors is unknown.

@@

\section{Methods}
\subsection{Study Area}
The study area was centered at -17\degree 13'9''S, 23\degree 56'4''E
(\Cref{StudyArea}a) and was represented by a rectangular bounding box that
stretched over 1.3 Mio. km\textsuperscript{2}, ecompassing the entire KAZA-TFCA
(\Cref{StudyArea}b). The KAZA-TFCA is the world's largest transboundary
conservation area and comprises parts of Angola, Botswana, Namibia, Zimbabwe,
and Zambia, covering a total of 520'000 km\textsuperscript{2}. Its landscape
varies regionally and ranges from savanna, to grassland and from dry to moist
woodland habitats. A dominant hydrogeographical feature in its center is the
flood-pulsing Okavango Delta. The delta and its surroundings are considered a
stronghold for African wild dogs that may act as a source for the recolonization
of surrounding habitats \citep{Cozzi.2013}. The wet season within the KAZA-TFCA
lasts from November to March and is out of phase with the main flooding of the
Okavango Delta which peaks between July and August \citep{McNutt.1996,
Wolski.2017}. While large portions of land within the KAZA-TFCA are designated
national parks or other protected areas, substantial human influence remains due
to roads, agricultural sites and settlements and villages.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_StudyArea.png}
    \caption{(a) The study area of our research was confined by a bounding box
    encompassing the entire KAZA-TFCA, which comprises parts of Angola, Namibia,
    Botswana, Zimbabwe, and Zambia. (b) The KAZA-TFCA represents the world's
    largest terrestrial conservation area and covers a total of 520'000
    km\textsuperscript{2}. Its purpose is to re-establish connectivity between
    already-existing national parks and other protected areas (green polygons).
    The dispersal data used in this study was collected on a free-ranging
    African wild dog population inhabiting the Moremi National Park in northern
    Botswana.}
    \label{StudyArea}
  \end{center}
\end{figure}

\subsection{GPS Relocation Data}
Between 2011 and 2019, we collected GPS relocation data on dispersing wild dogs
from a free-ranging wild dog population inhabiting the Moremi National Park in
northern Botswana. We selected potential dispersers based on age, pack size,
number of same‐sex siblings within the pack, and presence of unrelated
opposite-sex individuals in the pack \citep{McNutt.1996, Behr.2020}. We
immobilized selected individuals using a cocktail of Ketamine/Xylazine/Atropine
\citep{Osofsky.1996, Cozzi.2020} that was injected by dart, fired from a
CO\textsubscript{2}-pressurized gun (\textit{DAN-Inject, Denmark}). Anesthesia
protocols were approved by the Ministry of Environment, Natural Resources
Conservation and Tourism of Botswana (permit EWT 8/36/4 XXXVI). Immobilized
individuals were fitted with GPS/Satellite radio collars (\textit{Vertex Lite;
Vectronic Aerospace GmbH, Berlin}) that guaranteed automated drop-off through a
decomposable piece of cotton. Handling and collaring of all individuals was
supervised by a Botswana-registered wildlife veterinarian and all individuals
rejoined their pack within one hour after immobilization. 16 collared
individuals eventually dispersed, each in a separate same-sex coalition (7
female and 9 male coalitions).

During dispersal, collars were programmed to record a GPS fix every 4 hours.
Collected relocations were regularly transmitted over the Iridium satellite
system, which allowed remote tracking of individuals, even if they left the main
study area and crossed international borders. In some instances exact dispersal
dates were known from field observations. In other cases, we distinguished
between residency and dispersal using the net-squared displacement metric. This
metric measures the squared Euclidean distance a GPS relocation to a reference
point \citep{Borger.2012}, which in our case was set to the center of each
individual's natal home range. As such, dispersal was deemed to have started
when an individual left its natal home range and ended once individuals became
sedentary again. Because behavior during dispersal is more pertinent for
assessing connectivity, we discarded all data that was collected during
residency and only retained GPS data recorded during dispersal. As previous
research revealed similar behavior of females and males during dispersal
\citep{Woodroffe.2019, Cozzi.2020}, we did not distinguish between sexes. After
collection, we converted collected GPS coordinates (n = 4'169) to steps, where
each step represented the straight-line distance traveled by and individual
between two consecutive GPS relocations \citep{Turchin.1998}. To ensure a
regular sampling interval, we removed fixes that were not collected on the
4-hourly schedule.

\subsection{Covariates}
We represented the physical landscape across the study area using a set of
habitat covariates including water-cover, distance to water, tree-cover, and
shrub/grassland-cover. Because water cover greatly changes within and between
years around the Okavango Delta, we applied a remote sensing technique that
allowed us to generate frequently updated water cover layers and corresponding
distance to water layers for the close surroundings of the Delta (see
\citealp{Wolski.2017} and \citealp{Hofmann.2021}). Thus, those layers temporally
aligned with our dispersal events. We also computed a proxy for human influence,
depicting anthropogenic pressures stemming from human-density, agricultural
sites, and roads. All obtained layers were coarsened or interpolated to a target
resolution of 250 m by 250 m and cropped slightly beyond the extent of the
KAZA-TFCA. Further details on the sources and preparation of each habitat
covariate are given in \cite{Hofmann.2021}.

Besides habitat covariates, we also computed movement metrics that we used as
movement covariates in our models. Movement metrics were calculated for each
step and included the step length (\textsf{sl}), its natural logarithm
(\textsf{log(sl)}), and the cosine of the relative turning angle
(\textsf{cos(ta)}). Because wild dogs follow a diurnal activity pattern, we also
coded a binary variable (\textsf{LowActivity}) indicating whether a step was
realized during periods of low wild dog activity (17:00 to 07:00 local time) or
high wild dogactivity (09:00 to 17:00 local time). Handling and manipulation of
all data, as well as all modelling and simulating was achieved with statistical
software {\tt R}, version 3.6.6 \citep{R.2019}. {\tt R}-scripts to reproduce the
analyses are provided on Github.

\subsection{Movement Model}
We used integrated step selection functions (iSSF) to parametrize a mechanistic
movement model of dispersing wild dogs. In contrast to regular step selection
functions, \textit{integrated} step selection functions allow simultaneous
inference on movement and habitat preferences of the studied animal, as well as
on potential interactions between the two. This results in less biased selection
estimates and allows to treat the resulting model as a fully mechanistic
movement model based on which movement can be simulated \citep{Avgar.2016}. To
conduct iSSF analysis, we paired each observed step with 24 random steps.
Together, an observed and its random steps formed a stratum that received a
unique identifier. The 24 random steps served as pseudeo-absence and represented
a potential alternative that the animal could have realized but decided not to.
We generated random steps by sampling random turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from a gamma distirbution that
was fitted to observed steps (scale = 6'308, shape = 0.37). Although a larger
number of random steps reduces the sampling error \citep{Avgar.2016},
preliminary analyses revealed that 24 steps were sufficient in our case. Along
each step, we extracted and averaged spatial covariates using the {\tt velox}
package and we calculated the movement metrics \textsf{sl}, \textsf{log(sl)},
and \textsf{cos(ta)}. To facilitate model convergence, we standardized all
continuous covariates to a mean of zero and a standard deviation of one. Since
correlation among covariates was low (\(|r| > 0.6\); \citealp{Latham.2011}), we
retained all of them for modeling.

To contrast realized steps (scored 1) and random steps (scored 2) and fit a
movement model, we assumed that animals assigned a selection score \(w(x)\) to
each step. The score depended on the step's covariates \(X\), as well as the
animals preferences \(\beta\) towards these covariates. The probability of a
step being realized was then contingent on the step's selection score, as well
as on the selection scores of all other step in the same stratum. To estimate
the preferences of interest, we ran conditional logistic regression models in
the r-package {\tt glmmTMB}. More specifically, we applied the method developed
by \citep{Muff.2020}, which allows to fit random slopes. Here, we treated animal
IDs as random effect and modelled random slopes for each covariate and we fixed
the random intercept variance at an arbitrary high value of
10\textsuperscript{6} to make use of the  ``poission''-trick  \citep{Muff.2020}.

The movement model was based on a habitat selection model for dispersing wild
dogs presented in \cite{Hofmann.2021}. In the original model, no interactions
between the habitat and movement covariates were considered, so we slightly
expanded this base model by proposing interactions between movement and habitat
covariates. More specifically, we started with the base model and incrementally
increased model complexity by adding all possible two-way interactions between
habitat covariates and movement covariates. For instance, for the covariate
\textsf{water}, we proposed the interactions \textsf{Water:log(sl)},
\textsf{Water:log(sl)}, and \textsf{Water:cos(ta)}. Besides those combinations,
we also proposed the interactions \textsf{sl:cos(ta)} and
\textsf{log(sl):cos(ta)} to account for a correlation between turning angles and
step lengths, as well as the interactions \textsf{sl:LowActivity} and
\textsf{log(sl):LowActivity} to account for the fact that step lengths may
differ due to wild dogs' diurnal activity pattern. To compare competing models
and assess the most parsimonious movement model, we ran stepwise forward model
selection based on Akaike's Information Criterion (AIC, \citealp{Burnham.2002}).

We validated the predictive power of the most parsimonious movement model using
k-fold cross-validation for case-control studies as presented
\cite{Fortin.2009}. We randomly split the input data into training and testing
data with an 80:20 ratio. Using the training data we parametrized a movement
model based on which we predicted selection scores \(w(x)\) for all steps in the
test data. Within each stratum we then assigned ranks 1-25 to each step based on
predicted selection scores such that rank 1 was given to the step with the
highest score \(w(x)\). Across all strata we determined the realized step's rank
and calculated rank frequencies for realized steps. Finally, we computed
Spearman's rank correlation between ranks and associated frequencies \(r_{s,
realized}\). We replicated the entire procedure 100 times and computed the mean
correlation coefficient (\(\bar{r}_{s, realized}\)), as well as its 95\%
confidence interval across all replicates. For comparison, we repeated the same
procedure 100 times assuming random preferences. Random preferences were
implemented by discarding the realized step from all strata and identifying the
rank of a random step in each stratum. Again, we calculated Spearman's rank
correlation coefficient (\(r_{s, random}\)), its mean across repetitions
(\(\bar{r}_{s, random}\)), and its 95\% confidence interval. This validation
proofs a significant prediction in case the confidence intervals of
\(\bar{r}_{s, realized}\) and \(\bar{r}_{s, random}\) do not overlap.

\subsection{Dispersal Simulation}
We used the most parsimonious movement model to simulate 80'000 dispersing wild
dogs moving across the KAZA-TFCA. The simulation resembled an inverted iSSF and
worked as follows. (1) We defined a random source point and assumed a random
initial orientation of the animal. (2) Departing from the source point, we
generated 25 random steps by sampling relative turning angles from a uniform
distribution (\(-\pi, +\pi\)) and step lengths from our fitted gamma
distribution. As such, each step corresponded to the straight line movement
conducted within 4 hours. To prevent unreasonably large steps, we capped sampled
values to a maximum of 35 km, which corresponds to the farthest distance ever
traveled within 4 hours by on of our dispersers. (3) Along each random step, we
extracted and averaged habitat covariates and calculated movement covariates.
Extracted values were again standardized using the same transformations that
were used on the input data. (4) We applied the parameterized movement model and
predicted the selection score \(w(x)\) for each of the proposed steps and
translated selection scores into probabilities by applying \Cref{EQ2}. (5) We
randomly sampled one of the random steps based on predicted probabilities and
used it to determine the animal's new position. We repeated steps (2) to (5)
until a total of 2'000 steps were realized.

To minimize the influence of edge effects and to deal with random steps leaving
the study area, we followed \citep{Koen.2010} and artificially expanded all
covariate layers by adding a 100 km buffer zone. Within the buffer zone, we
randomized covariate values by resampling values from the original covariate
layers. In result, simulated dispersers were able to leave and reenter the main
study area through this buffer zone. In cases where a disperser approached the
outer map boundary such that one of the proposed random steps transgressed its
border, we resampled transgressing steps until they fully lied within the buffer
zone. Although the algorithm for the simulation was mainly coded in {\tt R}, we
wrote a few helper functions in {\tt C++} and imported them into {\tt R} using
the {\tt Rcpp} package \citep{Eddelbuettel.2011, Eddelbuettel.2013}.

\subsection{Source Points}
We initiated virtual dispersers at 50'000 randomly selected source points within
contiguous protected areas larger \(>\) 700 km\textsuperscript{2}
(\Cref{SourcePoints} (a)). This conforms to the average home range requirement
of resident wild dogs \citep{Pomilia.2015} and allowed us to remove areas too
small to host viable wild dog populations. By distributing source points
randomly, the number of source points per km\textsuperscript{2} within protected
areas was approximately equal across the study area. To render potential
immigrants into the study system, we placed additional 30'000 source points
inside the buffer zone around the main study area (\Cref{SourcePoints} (b)).
This resulted in a total of 80'000 source points, each representing the start
point of an individual disperser.

\begin{figure}[htbp]
  \begin{center}
    \includegraphics[width = \textwidth]{99_SourceAreas.png}
    \caption{(a) Different source areas from which we released virtual
    dispersers. We only considered contiguous protected areas (national parks,
    game reserves, and forest reserves) that were larger than 700
    km\textsuperscript{2} (green). This area corresponds to the average home
    range requirement for viable wild dog populations \citep{Pomilia.2015}. To
    render potential immigrants into the study system, we also initiated
    dispersers within a buffer zone (blue) surrounding the main study area. (b)
    Source points from which dipsersers were released. 50'000 dispersers were
    released from the main study area (green dots) and another 30'000 dispersers
    within the virtual buffer (blue dots).}
    \label{SourcePoints}
  \end{center}
\end{figure}

\subsection{Heatmap}
To identify dispersal hotspots across our study area using the simulated
dispersal events, we rasterized all simulated trajectories and created a heatmap
indicating the absolute frequency at which each raster-cell in the study area
was visited by our virtual dispersers. If the same trajectory crossed a
raster-cell twice, it was only counted once, meaning that we did not consider
revisits. This mitigated biases arising due to trapped individuals that were
moving in circles. To achieve high performance rasterization of all simulated
trajectories, we used the R-package {\tt terra} \citep{Hijmans.2020}.
